Copulas and their potential for ecology

### 2019-10-06 Paper.Rmd
added TL stuff to Q3 sections

We next addressed Q3 by exploring, using both data and models, three hypothesized
ecological consequences of non-normal copula structure and asymmetric tail
associations. As for Q2, models used are intended as initial explorations, only,
of whether copula structure may possibly have a meaningful influence on important ecological
phenomena. Therefore only simple models are used, and comprehensive explorations of the  
sensitivity of results to model strucuture and parameters are left for future work.

@@

To help answer Q3, we tested the above hypothesis
using our multivariate datasets (Table \ref{tab_data_info}).
For each dataset we calculated the spatial mean time series, and then 
the skewness through time of that
mean. Then, for each dataset, we compared the value obtained 
to a distribution of values 
of the same quantity for each of 10000 surrogate datasets. Surrogate 
datasets were produced by randomizing the empirical data
in a special way to have the 
copula structure of a multivariate-normal distribution, but 
to retain exactly the same
distributions of values for each sampling location as the original data
(Appendix \ref{SM-Surrog_multi}). 
Surrogates also had very similar Spearman correlations 
between pairs of sampling locations
as the data. 
Our comparisons therefore tested the null 
hypothesis that the skewness of the spatial 
mean took values on the empirical data no different 
from what one would expect if the copula
structure of the data were multivariate normal (i.e., the same copula as a 
multivariate normal distribtion), but the data were otherwise statistically
unchanged. Significant differences indicate that non-normal copula structure in the data 
contributed to the skewness of the spatial mean time series, 
i.e., to its instability and "spikiness" through time.

@@

Pursuing ideas that sprang from earlier theoretical work [@Reuman2017], we 
tested whether the copula structure of the dependences between population variables 
measured in different locations influences the spatial version of Taylor's
law. Taylor's law is a commonly observed and widely studied 
[@Taylor1961; @Taylor1988; @Cohen2013; @Xu2016; @Reuman2017] 
empirical pattern that relates the 
variances of groups, $g$, of population measurements to the means of the 
groups via a power law, $v_g = a \times m_g^b$, or equivalently, 
$\log(v_g)=\log(a)+b \times \log(m_g)$. Here $b$ is called 
the Taylor's law *exponent* or 
*slope*, and $\log(a)$ is the Taylor's law intercept. 
There are several versions of Taylor's law. For spatial Taylor's law,
given population density or abundance data $x_i(t)$ measured in locations
$i=1,\ldots,N$ and times $t=1,\ldots,T$, a group $g$ consists of 
all the measurements $x_i(t)$, $i=1,\ldots,N$ made in different locations at the same time.
So means $m_g$ and variances $v_g$ are computed across space. We consider that 
Taylor's law holds true for a dataset if the 
$\log(v_g)$ versus $\log(m_g)$ scatter plot 
for the dataset shows a linear, homoskedastic pattern.
Taylor's law has been 
verified empirically for hundreds of taxa and has been applied in numerous fields
including fisheries management, estimation of species persistence times, and
agriculture [@Cohen2015; @Reuman2017].

<!--***DAN: continung to add TL stuff here, and then next in the following results section-->

@@

<!--***DAN: the below is on TL and needs to be augmented if we keep TL results-->
We also substantiated using models (Appendix \ref{SM-Consequence_TL_simulation}, \ref{SM-sect:copsurrognd}, Fig. \ref{SM-fig_spTL})
a hypothesis (Fig. \ref{MasterFigure}, arrow labelled "Z") that 
Taylor's law can be influenced by
tail dependence.

@@

>>>>

We addressed Q3 by exploring, using both data and models, three hypothesized
ecological consequences of non-normal copula structure and asymmetric tail
associations. As for Q2, models used are intended as initial explorations, only,
of whether copula structure may possibly have a meaningful influence on important ecological
phenomena. Therefore only simple models are used, and comprehensive explorations of the 
sensitivity of results to model strucuture and parameters are left for future work.

@@

We tested the above hypothesis
using our multivariate datasets (Table \ref{tab_data_info}).
For each dataset we calculated the spatial mean time series, and then 
the skewness through time of that
mean. Then, for each dataset, we compared the value obtained 
to a distribution of values 
of the same quantity for each of 10000 surrogate datasets. Surrogate 
datasets were produced by randomizing the empirical data
in a special way to have the 
copula structure of a multivariate-normal distribution, but 
to retain exactly the same
distributions of values for each sampling location as the original data
(Appendix \ref{SM-Surrog_multi}). 
Surrogates also had very similar Spearman correlations 
between pairs of sampling locations
as the data. 
Our comparisons therefore tested the null 
hypothesis that the skewness of the spatial 
mean took values on the empirical data no different 
from what one would expect if the copula
structure of the data were multivariate normal (i.e., the same copula as a 
multivariate normal distribtion), but the data were otherwise statistically
unchanged. Significant differences indicate that non-normal copula structure in the data 
contributed to the skewness of the spatial mean time series, 
i.e., to its instability and "spikiness" through time.

@@

Finally, and pursuing ideas that sprang from earlier theoretical work [@Reuman2017], we 
tested whether the copula structure of the dependences between population variables 
measured in different locations has consequences for the spatial version of Taylor's
law. Taylor's law is a commonly observed and widely studied 
[@Taylor1961; @Taylor1988; @Cohen2013; @Xu2016; @Reuman2017] 
empirical pattern that relates the 
variances of groups, $g$, of population measurements to the means of the 
groups via a power law, $v_g = a \times m_g^b$, or equivalently, 
$\log(v_g)=\log(a)+b \times \log(m_g)$. Here $b$ is called 
the Taylor's law *exponent* or 
*slope*, and $\log(a)$ is the Taylor's law intercept. 
There are several versions of Taylor's law. For spatial Taylor's law,
given population density or abundance data $x_i(t)$ measured in locations
$i=1,\ldots,N$ and times $t=1,\ldots,T$, a group $g$ consists of 
all the measurements $x_i(t)$, $i=1,\ldots,N$ made in different locations at the same time.
So means $m_g$ and variances $v_g$ are computed across space. We refer to the matrix
with $x_i(t)$ in the $t$th row and the $i$th column as a *population matrix*. 
We consider that 
Taylor's law holds true for a dataset if the 
$\log(v_g)$ versus $\log(m_g)$ scatter plot 
for the dataset shows a linear, homoskedastic pattern.
Taylor's law has been 
verified empirically for hundreds of taxa and has been applied in numerous fields
including fisheries management, estimation of species persistence times, and
agriculture [@Cohen2015; @Reuman2017], so the question of whether copula
structure of spatial dependence influences Taylor's law may have applied
significance.

To explore the influence of copula structure and tail associations on spatial Taylor's
law, we carried out a series of simulations that generated sets of population 
matrices that had 
distinct copula structure between locations but that 
were otherwise statistically similar. For $k=1,\ldots,1000$
and X representing a Clayton, normal, Frank, or a survival Clayton copula family,
we generated $N \times n$ population matrices $m^{(\text{X},k)}$ with the following properties:
1) for any given values of $j$ and $k$, the values 
$\{ m_{t,j}^{(\text{X},k)} : t=1,\ldots,N  \}$ were exactly the same, as unordered
sets, for all X, i.e., the same actual population values were used for location $j$ 
in simulation $k$, regardless of X;
2) for any given $k$, $j_1 \neq j_2$, the Spearman correlations (computed through time)
$\cor_t(m_{t,j_1}^{(\text{X},k)},m_{t,j_2}^{(\text{X},k)})$ were the same, to within sampling
variation, i.e., correlations through time of populations in two locations were the same,
up to sampling variation, regardless of X;
3) the copula for the dependence between $m_{t,j_1}^{(\text{X},k)}$ and $m_{t,j_2}^{(\text{X},k)}$ was X.
We used $N=50$ and $n=25$. Details of how these matrices were generated are in 
Appendices \ref{SM-Consequence_TL_simulation} and \ref{SM-sect:copsurrognd}.
Thus for each $k$, the poulation matrices $m^{(\text{X},k)}$
for X taking the values Clayton, normal, Frank, and survival Clayton
were statistically similar except for the copula structure between locations, which was
X. Thus comparing how Taylor's law may manifest differently for $m^{(\text{X},k)}$
for different values of X constitutes a test of the influence of copula
structure on Taylor's law. We selected X equal to the Clayton, normal, Frank, and survival 
Clayton copulas to explore a range of tail association patterns.

A variety of Taylor's law statistics were computed for each populaton matrix
$m^{(\text{X},k)}$. First, the $N=50$ 
spatial means,
$m_g$, and variances, $v_g$, were computed, and $\log(v_g)$ vs. $\log(m_g)$ plots 
were considered. Linearity of the $\log(v_g)$ vs. $\log(m_g)$ relationship was 
tested for each simulation by comparing the linear regression
through these $50$ points to a quadratic alternative via an $F$-test, 
producing a $p$-value result. If 
these $p$-values were uniformly distributed across the unit interval for the 1000 replicate 
simulations which were generated for X, it supported the 
linearity assumption of Taylor's law for X; whereas if 
they were clustered toward smaller values
the test tended to reject that assumption. We also tested, for each simulation, 
the assumption of Taylor's 
law that the $\log(v_g)$ vs. $\log(m_g)$ plot was homoskedastic: we regressed the absolute 
residuals of the 
linear regression of $\log(v_g)$ versus $\log(m_g)$ against the predictions of that 
regression. A significant $p$-value
result of this test indicates heteroskedasticity. For each simulation we also computed the
root mean squared error of $\log(v_g)$ vs. $\log(m_g)$ data from the linear regression
of $\log(v_g)$ versus $\log(m_g)$, as well as the 
intercept and slope of the linear regression.
Finally we recorded the quadratic coefficient of the regression of $\log(v_g)$ against
$\log(m_g)$ and $(\log(m_g))^2$, and the mean curvature of 
the quadratic regression equation itself
across the values $\log(m_g)$. The distrubutions of all these 
statistics across the 1000 replicate 
simulations were compared for X the Clayton, normal, Frank, and survival Clayton copulas, 
to determine if copula structure influences Taylor's law.

@@

Also consistent with hypothesis, copula structure had a substantial effect on
Taylor's law. 
Taylor's law was strongly influenced, and was often even invalidated in its assumptions of 
linearity and homoskedasticity, by non-normal copula structure. For normal copula structure 
(i.e., for simulations that gave normal-copula dependence between populatons
in different locations), $p$-values for the linearty
and homoskedasticity tests were roughly uniformly distributed across replicate 
simulations and Taylor's law appeared visually to be a reasonable approximation
of the $\log(v_g)$ versus $\log(m_g)$ relationship (Fig. \ref{fig_spTL}B,E,F). 
Furthermore, quadratic coefficients
and curvature values were close to $0$ (Fig. \ref{fig_spTL}J,K), and root mean squared
errors from the linear regression were relatively small (Fig. \ref{fig_spTL}G).
But linearity or homoskedasticity were violated more frequently for non-normal 
copula structure
(Fig. \ref{fig_spTL}A,C,D,E,F); quadratic coefficients and curvatures were frequently 
non-zero (Fig. \ref{fig_spTL}J,K); and root mean squared
errors from the linear regression were much higher (Fig. \ref{fig_spTL}G). Slopes 
and intercepts of the linear regression were also strongly affected by copula
structure (Fig. \ref{fig_spTL}H,I), though some of the effect here was because
linear regressions do not always adequately represent the $\log(v_g)$ versus $\log(m_g)$
relationship when copula structure was not normal.  
Thus our results substantiated the hypothesis, at least for the models we used,
that Taylor's law can be influenced by copula structure and asymmetric
tail associations. This is represented in Fig. \ref{MasterFigure} by the 
arrow labelled "Z".

@@

<!--adding plots showing influences of non-normal copula on spatial Taylors law-->
\begin{figure}[!h]
\textbf{ \hspace{1.4cm} (A) \hspace{2.92 cm} (B) \hspace{2.92 cm} (C) \hspace{2.92 cm} (D)} \\
\vspace{-0.5 cm} 
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_CNFSC_spearman.png}\\
\end{center}
\textbf{ \hspace{0.6 cm} (E) \hspace{1.4 cm} (F) \hspace{1.4 cm} (G) \hspace{1.4 cm} (H) \hspace{1.4 cm} (I) \hspace{1.4 cm} (J) \hspace{1.4 cm} (K)}\\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=16 cm]{./Results/taylorslaw_results/SpatialTL_7stats_spearman.pdf}
\caption[Spatial Taylor's law results]{Spatial Taylor's law results. Log(spatial variance) vs. 
log(spatial mean) relationships for $1000$ simulations over $50$ time steps for $25$ locations 
using Clayton (A), normal (B), Frank (C), and survival Clayton (D) copula structures between
locations. Results from all $1000$ 
simulations were plotted on the same axes for each panel, but two example simulations are 
shown in red and green to help assess to what extent variation on the plots was between 
simulations or within simulations. For each simulation, we tested the linearity 
(E) and homoskedasticity (F) of the log(variance) vs. log(mean) plot for that simulation, 
and quantified the root mean squared error of points from the linear regression line (G).
We also calculated the linear regression intercept (H) and slope (I), the quadratic
term of a quadratic regression through the points on the log(variance) vs. log(mean) plot
(J), and the curvature of that quadratic regression (K). Distributions of values 
across all 1000 simulations are displayed. Spearman-preserving surrogates were used, 
though results using Kendall-preserving surrogates were similar. See the methods
section for Q3 and Appendices \ref{SM-Consequence_TL_simulation} and 
\ref{SM-sect:copsurrognd} for details.\label{fig_spTL}}
\end{center}
\end{figure}
<!--***DAN: Shyamolina, the spacing of the panel labels (A, B, C, etc.) is not quite 
matched up with the spacing of the panels themselves, can you fix it please?-->
